---
title: "Tract-Level Entry-Exit Analysis, Philadelphia"
author: "Jesse Cahill"
date: "January 24, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load packages
library(here)
library(tidyverse)
library(viridis)
source(here("src", "get_colnames.R"))
```

# Introduction
**Regarding Intervals**:
  The system of using intervals to describe when businesses come into and go out of business can be inherently confusing.  Some notes:
  - The NETS data for a year is based on a survey taken in January of that year.  So, if a business appears for the first time in 2002, it ACTUALLY showed up some time in 2001 and was not seen until the next survey.  Entry is one year behind NETS records.
  - If a business is seen for the last time in 2009, that means it was present in January of 2009 but not January of 2010.  Therefore, exit is concurrent with NETS records. 
  - When we plot The way entryies and exits will be formatted moving forward is left aligned, they'll show up for the first

# Read, Tidy
To form an effective baseline, we'll look at entry/exit rates for all businesses using a 1 year interval. We'll add a variable for net business change (entry - exit) for each year, tract, and business type. We'll try analysis on both net change and entry/exit seprately.

```{r}
yr1_path <- here("data", "final", "nets_philly_entryExit_1yrInterval.csv")
col_names <- get_colnames(yr1_path)
data <- read_csv(yr1_path, skip=3, col_names = col_names)
```


```{r}
# net and enter/exit data separated due to index differences
tidy_net <- data %>% 
  gather("key", "value", ends_with("enter_year"), ends_with("exit_year")) %>% 
  separate(key, into=c("business_type", "year_type"), sep="\\.") %>% 
  mutate(business_type = as_factor(business_type)) %>% 
  spread(year_type, value) %>% 
  mutate(net = enter_year - exit_year) %>% 
  select(t10_cen_uid_u_2010, year, business_type, net)
  
tidy_entex <- data %>% 
  gather("key", "value", ends_with("enter_year"), ends_with("exit_year")) %>% 
  separate(key, into=c("business_type", "year_type"), sep="\\.") %>% 
  mutate(business_type = as_factor(business_type)) %>% 
  spread(year_type, value) %>% 
  gather("behavior", "num_businesses", enter_year, exit_year)

```

# Baseline Analysis - 1 Year Intervals, Total business count

```{r}
total_net <- tidy_net %>% 
  filter(business_type == "TOTAL")

total_entex <- tidy_entex %>% 
  filter(business_type == "TOTAL")
```


## Time Series Line Graphs

### Aggregated by Year

```{r}

net_time <- ggplot(total_net, aes(
  x = year,
  y = net
)) + 
  stat_summary(fun.y = "mean", geom="line") +
  ggtitle("Mean Tract-Level Net Business Change, Philadelphia")

entex_time <- ggplot(total_entex, aes(
  x = year,
  y = num_businesses,
  color = behavior
)) + 
  stat_summary(fun.y = "mean", geom="line") +
  ggtitle("Mean Tract-Level Enter/Exit Business Rates, Philadelphia")


net_time
entex_time
```

This is interesting due to things we've seen previously showing that nearly every business type shows a dip in total number of US businesses in 2010, with a subsequent recovery in 2011. In Philly overall business growth dropped from a net *+*~27 per tract in 2008 to ~*-*25 per tract in 2009.  By 2010, which historically we believed was the low point of the Great Recession in terms of business numbers, net change of businesses is already recovered to +10. After thinking back more, I realized that this is likely due to the way the NETS is structured, taking a snapshot in January.  The dip in businesses in 2010 actually occured in 2009 but wasn't shown until January of 2010.

We can add further granularity by looking at entry and exit rates specifically. Exit rates have a huge increase between 2008 and 2009, but go back nearly to normal again in 2010. Entry rates experience an unremarkable dip between 2008 and 2009, then see a huge spike in 2010.

## Continuous Categorical Stuff
**look at these again in terms of percent change from original values for better comparison**

### Net Change

```{r}
#factor out total, it'll just throw off the scale
net_heat <- filter(tidy_net, business_type != "TOTAL") %>% 
  group_by(year, business_type) %>% 
  summarize(net = mean(net)) %>% 
  ungroup() %>% 
  filter(!is.na(net))

theme_heat <- theme_classic() +
  theme(axis.line = element_blank(),
        axis.ticks = element_blank())

plot <- ggplot(net_heat, aes(x = year, y = business_type)) +
  geom_tile(aes(fill = net), color = "white") +
  coord_fixed() + theme_heat + 
  scale_fill_distiller(palette = "RdBu") +
  ggtitle("Time-Series Net Business Change Rate")

plot
```

There's some big differences in color between WALH and the other two categories hehre, which makes it kind of difficult to discern differences between them.  Its difficult to discern anything from FFAH.  ACTH shows a noticible net decrease in 2009, and then a recovery in 2010.  2001 also seems to be a big year of growsh for both ACTH and WALH.   
In 2001 we see some pretty drastic growth of WALH, and then a pretty big dropoff in 2004.  Relating to the recession, we see a net positive growth in 2007 which softens in 2008 and then bottoms out in 2009, only to rebound dbig in 2010.

## Enter/Exit Change

```{r}
#factor out total, it'll just throw off the scale
entex_heat <- filter(tidy_entex, business_type != "TOTAL") %>% 
  group_by(year, business_type, behavior) %>% 
  summarize(num_businesses = mean(num_businesses)) %>% 
  ungroup() %>% 
  filter(!is.na(num_businesses))


plot1 <- ggplot(entex_heat, aes(x = year, y = business_type)) +
  geom_tile(aes(fill = num_businesses), color = "white") +
  coord_fixed() +
  facet_grid(rows = vars(behavior)) +
  theme_heat +
  scale_fill_viridis() +
  ggtitle("Time-Series Enter-Exit Rates")

plot1
```

In hindsight this isn't the best way to look at things, we'll just plot timeseries instead.


```{r, fig.width=10, fig.height=5}

entex_timeseries <- ggplot(entex_heat, aes(x = year, y = num_businesses)) +
  geom_line(aes(color = business_type)) +
  coord_fixed() +
  facet_grid(rows = vars(behavior)) +
  ggtitle("Time-Series Enter-Exit Rates")

net_timeseries <- ggplot(net_heat, aes(x = year, y = net)) +
  geom_line(aes(color = business_type)) +
  coord_fixed() +
  ggtitle("Time-Series Net Change Rates")

entex_timeseries
net_timeseries
```


## Spatial maps
**Shiny app???**
```{r}
library(sf)
library(sp)

philly_cbsa_shp <- here("gis", "philly_cbsa_tract_2010_merge")
philly_cbsa_geo <- read_sf(philly_cbsa_shp)

cbsa_geo_enterexit <- philly_cbsa_geo %>% 
  inner_join(
    mutate(tidy_entex, GEOID10 = as.character(t10_cen_uid_u_2010)))

cbsa_geo_enterexit %>% 
  filter(year==2010, business_type=="TOTAL") %>% 
  ggplot() +
  geom_sf(aes(fill=num_businesses)) +
  facet_wrap(~behavior) +
  scale_fill_viridis()


```

Tract level data is too granular for this big of an area, some center-city tracts with huge numbers are throwing off the scale of the whole thing. And even if the scale was right, the tracts in urban areas are so comparitively small we can't even see what their values are. We'll have to aggregate to a bigger geographic unit later on. 

### Tract Level, City of Philadelphia
If we zoom to a smaller unit of area we can examine things there on the tract level.

```{r}
philly_city_shp <- here("gis", "philly_tract_2010_merge")
philly_city_geo <- read_sf(philly_city_shp) %>% 
  st_transform(4326, check = TRUE)

city_geo_enterexit <- philly_city_geo %>% 
  inner_join(
    mutate(tidy_entex, GEOID10 = as.character(t10_cen_uid_u_2010)) %>% 
      select(-t10_cen_uid_u_2010)) %>% 
  spread(behavior, num_businesses) %>% 
  inner_join(
    mutate(tidy_net, GEOID10 = as.character(t10_cen_uid_u_2010)) %>% 
      select(-t10_cen_uid_u_2010)) 


```
```{r}
#write this to file for shiny
st_write(city_geo_enterexit, here("philly_enter_exit", "data.shp"))
```

Those few outliers are still making things a bit difficult for us. We see that often the same tracts show up brightly in the heatmaps of all maps, whether looking at entry rates, exit rates, or net change.  There are a few exceptions with high entry rates and net change in the northern part of the city, but these are largely centered around center city.  We have to impose some common scale on these data if we want to see how they change relative to themselves.


### Interactive Maps

```{r}
write_csv(city_geo_enterexit)
```

```{r}
library(leaflet)

# m <- leaflet() %>% setView(lng = -75.1661, lat = 39.9534, zoom = 11)
# m %>% addTiles()
total_2010 <- city_geo_enterexit %>% 
  filter(year==2010, business_type=="TOTAL")

m <- leaflet(total_2010) %>%
  setView(lng = -75.1125, lat = 39.9971, zoom = 10.5) %>%
  addProviderTiles("MapBox", options = providerTileOptions(
    id = "mapbox.light",
    accessToken = Sys.getenv('MAPBOX_ACCESS_TOKEN')))

bins <- c(0, 10, 20, 50, 100, 200, 500, 1000, Inf)
pal <- colorBin("viridis", domain = total_2010$num_businesses, bins = bins)

m %>%
  addTiles() %>% 
  addPolygons(weight=1,
              color="white",
              fillColor = ~pal(num_businesses),
              opacity=.2,
              fillOpacity = .5) %>%
  addLegend(pal = pal,
            values = ~num_businesses,
            opacity = 0.7,
            title = NULL,
  position = "bottomright")
```

